Richness model

### 2024-08-16 julianna_coral_macroalg_wrangling.Rmd

```{r join}
macroAlgDiv %>% 
  full_join(betaDHabYr) %>% 
  # get just the columns we need
  select(-site_habitat, -hab_year, -island_side_hab_year) -> quadData

@@

```{r save}
 write_csv(quadData, here("data/macroalgae_diversity_quadrat.csv"))
```

@@

```{r get averages}
quadData %>% 
  group_by(year, site, habitat, island_side) %>% 
  summarize(N = n(),
            coral_cover_mean = mean(coral_cover),
            coral_cover_se = sd(coral_cover)/sqrt(N),
            coral_cover_hard_mean = mean(coral_cover_hard, na.rm = TRUE),
            coral_cover_hard_se = sd(coral_cover_hard, na.rm = TRUE)/sqrt(N),
            macroalgae_cover_mean = mean(macroalgae_cover),
            macroalgae_cover_se = sd(macroalgae_cover)/sqrt(N),
            macroalgae_cover_hard_mean = mean(macroalgae_cover_hard, na.rm = TRUE),
            macroalgae_cover_hard_se = sd(macroalgae_cover_hard, na.rm = TRUE)/sqrt(N),
            macroalgae_richness_mean = mean(macroalgae_richness),
            macroalgae_richness_se = sd(macroalgae_richness)/sqrt(N),
            macroalgae_shannon_mean = mean(macroalgae_shannon),
            macroalgae_shannon_se = sd(macroalgae_shannon)/sqrt(N),
            centroid_dist_habyear_mean = mean(hab_yr_distcentroid, na.rm = TRUE)
            ) -> siteData

@@

```{r save dataframe}
 write_csv(siteData, here("data/macroalgae_diversity_site.csv"))
```

@@

```{r calculate CV and summarize}
quadData %>% 
  # don't include year this time
  group_by(site, habitat, island_side) %>%
  summarize(N = n(),
            coral_cover_mean = mean(coral_cover),
            coral_cover_se = sd(coral_cover)/sqrt(N),
            coral_cover_hard_mean = mean(coral_cover_hard, na.rm = TRUE),
            coral_cover_hard_se = sd(coral_cover_hard, na.rm = TRUE)/sqrt(N),
            macroalgae_cover_mean = mean(macroalgae_cover),
            macroalgae_cover_se = sd(macroalgae_cover)/sqrt(N),
            macroalgae_cover_hard_mean = mean(macroalgae_cover_hard, na.rm = TRUE),
            macroalgae_cover_hard_se = sd(macroalgae_cover_hard, na.rm = TRUE)/sqrt(N),
            macroalgae_richness_mean = mean(macroalgae_richness),
            macroalgae_richness_se = sd(macroalgae_richness)/sqrt(N),
            macroalgae_shannon_mean = mean(macroalgae_shannon),
            macroalgae_shannon_se = sd(macroalgae_shannon)/sqrt(N),
            centroid_dist_habyear_mean = mean(hab_yr_distcentroid, na.rm = TRUE)
            )  -> alltimeCV
```

```{r save}
write_csv(alltimeCV, here(here("data/macroalgae_stability_metrics_all_time.csv")))
```

>>>>

```{r join}
macroAlgDiv %>% 
  full_join(betaDHabYr) %>% 
  rename(macroalgae_centroid_dist_habyear_mean = hab_yr_distcentroid) %>% 
  # get just the columns we need
  select(-site_habitat, -hab_year, -island_side_hab_year) -> quadData

@@

```{r save}
 write_csv(quadData, here("data/macroalgae_coral_diversity_quadrat.csv"))
```

@@

```{r get averages}
quadData %>% 
  group_by(year, site, habitat, island_side) %>% 
  summarize(N = n(),
            coral_cover_mean = mean(coral_cover),
            coral_cover_se = sd(coral_cover)/sqrt(N),
            coral_cover_hard_mean = mean(coral_cover_hard, na.rm = TRUE),
            coral_cover_hard_se = sd(coral_cover_hard, na.rm = TRUE)/sqrt(N),
            macroalgae_cover_mean = mean(macroalgae_cover),
            macroalgae_cover_se = sd(macroalgae_cover)/sqrt(N),
            macroalgae_cover_hard_mean = mean(macroalgae_cover_hard, na.rm = TRUE),
            macroalgae_cover_hard_se = sd(macroalgae_cover_hard, na.rm = TRUE)/sqrt(N),
            macroalgae_richness_mean = mean(macroalgae_richness),
            macroalgae_richness_se = sd(macroalgae_richness)/sqrt(N),
            macroalgae_shannon_mean = mean(macroalgae_shannon),
            macroalgae_shannon_se = sd(macroalgae_shannon)/sqrt(N),
            macroalgae_centroid_dist_habyear_mean = mean(macroalgae_centroid_dist_habyear_mean, na.rm = TRUE)
            ) -> siteData

@@

```{r save dataframe}
 write_csv(siteData, here("data/macroalgae_coral_diversity_site.csv"))
```

@@

```{r calculate CV and summarize}
quadData %>% 
  # don't include year this time
  group_by(site, habitat, island_side) %>%
  summarize(N = n(),

            # coral
            coral_cover_mean = mean(coral_cover),
            coral_cover_se = sd(coral_cover)/sqrt(N),
            coral_cover_hard_mean = mean(coral_cover_hard, na.rm = TRUE),
            coral_cover_hard_se = sd(coral_cover_hard, na.rm = TRUE)/sqrt(N),
            coral_cover_cv = CV(coral_cover),
            coral_cover_hard_cv = CV(coral_cover_hard),

            # macroalgae cover
            macroalgae_cover_mean = mean(macroalgae_cover),
            macroalgae_cover_se = sd(macroalgae_cover)/sqrt(N),
            macroalgae_cover_hard_mean = mean(macroalgae_cover_hard, na.rm = TRUE),
            macroalgae_cover_hard_se = sd(macroalgae_cover_hard, na.rm = TRUE)/sqrt(N),
            macroalgae_cover_cv = CV(macroalgae_cover),
            macroalgae_cover_hard_cv = CV(macroalgae_cover_hard),

            # macroalgae richness
            macroalgae_richness_mean = mean(macroalgae_richness),
            macroalgae_richness_se = sd(macroalgae_richness)/sqrt(N),
            macroalgae_shannon_mean = mean(macroalgae_shannon),
            macroalgae_shannon_se = sd(macroalgae_shannon)/sqrt(N),
            macroalgae_centroid_dist_habyear_mean = mean(macroalgae_centroid_dist_habyear_mean, na.rm = TRUE)
            )  -> alltimeCV
```

```{r save}
write_csv(alltimeCV, here(here("data/macroalgae_coral_diversity_stability_metrics_alltime.csv")))
```

### 2024-08-16 julianna_richness_model.Rmd

# load packages
require(here) # relative file paths
require(tidyverse) # data wrangling
require(lme4) # mixed models
```

@@

```{r load data}
cleaned_rich <- read_csv(here("data/algal_metrics_quadrat.csv")) # generated using my (Julianna's) algal metric code
```

# Make a model

```{r}
mRich <- glmer(,
               family = Poisson,
               data = cleaned_rich)
```

>>>>

# load packages
require(here) # relative file paths
require(tidyverse) # data wrangling
require(lme4) # mixed models

```

@@

```{r load data}
cleaned_rich <- read_csv(here("data/algae_diversity_quadrat.csv")) %>% # generated using my (Julianna's) algal metric wrangling code
  filter(year != 2020) %>% # fake data
  # make a unique quadrat ID
  mutate(quad_id = paste0(transect, "_", quadrat))

```

# Full model no site

WARNING: This is reeeeeally slow code to run

```{r make Poisson model}
mRich <- glmer(algae_richness ~ as.factor(year)*habitat + (1|site/quad_id),
               family = poisson,
               data = cleaned_rich)
```

# Examine output

```{r make prediction grid}
# create a grid to predict over
newdata <- expand_grid(year = unique(cleaned_rich$year),
                       habitat = unique(cleaned_rich$habitat),
                       site = unique(cleaned_rich$site),
                       quad_id = unique(cleaned_rich$quad_id))
```

```{r generate predictions}
newdata$Preds <- predict(mRich, newdata = newdata)
```

**IMPORTANT**: I don't think this is the correct way to get the error bars

```{r plot predictions}
newdata %>% 
  # get transect means
  group_by(year, site, habitat) %>% 
  summarize(Pred_mean_richness = mean(Preds),
            SD_richness = sd(Preds),
            N = n(),
            SE_richness = SD_richness/sqrt(N)
            ) %>% 
  ggplot(aes(x = year, y = Pred_mean_richness, group = habitat, color = habitat)) +
  geom_line() +
  theme_bw() +
  # add lines for disturbances
  geom_vline(xintercept = 2006, linetype = "dotdash") + # start of COTS
  geom_vline(xintercept = 2010,  color = "darkgray") + # cyclone Oli
  geom_vline(xintercept = 2019, linetype = "dashed", color = "red")  # bleaching

```

# Full model with site

```{r make Poisson model}
mRichSite <- glmer(algae_richness ~ as.factor(year)*habitat*site + (1|quad_id),
               family = poisson,
               data = cleaned_rich,
               control = glmerControl(optimizer = "bobyqa")) # this taken from : https://stackoverflow.com/questions/69929896/glmer-taking-days-to-compute-and-never-ending)
```

# Examine output

```{r make prediction grid}
# create a grid to predict over
newdataSite <- expand_grid(year = unique(cleaned_rich$year),
                       habitat = unique(cleaned_rich$habitat),
                       site = unique(cleaned_rich$site),
                       quad_id = unique(cleaned_rich$quad_id))
```

```{r generate predictions}
newdataSite$Preds <- predict(mRichSite, newdata = newdataSite)
```

```{r plot predictions}
newdataSite %>% 
  # get transect means
  group_by(year, habitat, site) %>% 
  summarize(Pred_mean_richness = mean(Preds),
            SD_richness = sd(Preds),
            N = n(),
            SE_richness = SD_richness/sqrt(N)
            ) %>% 
  ggplot(aes(x = year, y = Pred_mean_richness, group = habitat, color = habitat)) +
  geom_line() +
  theme_bw() +
  facet_wrap(~site) +
  geom_ribbon(ymin = Pred_mean_richness-(2*SE_richness),
              ymax = Pred_mean_richness+(2*SE_richness),
              alpha = 0.5) +
  # add lines for disturbances
  geom_vline(xintercept = 2006, linetype = "dotdash") + # start of COTS
  geom_vline(xintercept = 2010,  color = "darkgray") + # cyclone Oli
  geom_vline(xintercept = 2019, linetype = "dashed", color = "red")  # bleaching

```
